The Design and Implementation of the Synchronous Language \CEU

### 2016-03-29 paper.tex
...

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis to 
improve safety:
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements 
(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax 
(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes which can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
\CEU (b) for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events (including zero), depending on external stimuli the environment 
provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
external stimuli.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The execution of a program in \CEU is as follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to run in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Esterel and \CEU use discrete logical units of time in which the occurring 
events (which are the same in all boxes) have to fit somehow:
%
\begin{itemize}
\item \code{Box-1}:
Esterel with fixed-length ticks,
within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes inthe next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, the program can only detect one occurrence of \code{E}, which is 
considered in \code{R(D+E)}.
%
\item \code{Box-2}:
Esterel with variable-length ticks that adjust to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
start reaction \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{Box-3}:
\CEU with unique and queue-based policy for input events.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} simultaneous, 
which are handled in subsequent reactions \code{R(B)} and \code{R(C)}.
%
We assume that the CPU time for \code{R(B+C)} in Esterel is roughly the same as 
\code{R(B)+R(C)} in \CEU.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}

>>>>

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in fundamental 
semantic aspects.
%
Overall, \CEU has a simple semantics with fine-grained control for 
intra-reaction execution which is amenable to a simple temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class synchronized timers with a dedicated syntax.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing the practical aspect of our clean semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and device drivers.
%
We evaluated the expressiveness of \CEU in comparison to hand-crafted 
event-driven code in \C and attested a reduction in source code size (around 
25\%) with a small increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, applications have a bytecode footprint in the order of 
hundreds of bytes which can be transmitted over the air in a few 
packets~\cite{ceu.terra}.

@@

Figure~\ref{lst.abro} shows side-by-side the implementations in Esterel (a) and 
\CEU (b) for the following control specification:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''}~\cite{esterel.primer}.
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (ln. 3--8, in both 
implementations), i.e., Esterel's `$\|$' and \CEU's \code{par/and} constructs 
are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (ln. 2--9), which, in this case, serves the same purpose of 
\CEU's \code{par/or} (ln. 2--11):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

In the subsections that follow, we discuss the main differences between \CEU 
and Esterel:
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish with a summary of our design
    (\ref{sec.ceu.discussion}).

@@

Esterel defines time as a discrete sequence of logical unit instants or 
``ticks''.
At each tick, the program reacts to an arbitrary number of simultaneous input 
events, depending on external stimuli the environment provides.
In contrast, \CEU defines time as a discrete sequence of reactions to
unique input events.
At each input event, which constitutes a logical unit of time, the program 
reacts exclusively to it.
The execution of a program in \CEU is as follows~\cite{ceu.sensys13}:
%
\begin{enumerate}
\item The program initiates the ``boot reaction'' in a single trail (but 
      parallel constructs may create new trails).
\item Active trails execute until they await or terminate.
      This step is named a \emph{reaction chain}, and always runs in bounded 
      time.
\item The program goes idle and the environment takes control.
\item On the occurrence of a new external input event, the environment awakes 
      \emph{all} trails awaiting that event.
      It then goes to step 2.
\end{enumerate}
%
A program must react to an occurring event completely before handling the next.
%
Based on the synchronous hypothesis, a program takes no time on step 2 and is 
always idle on step 3.
In practice, if a new external input event occurs while a reaction chain is 
running (step 2), it is enqueued to occur in the next reaction.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in Esterel and 
\CEU.
The box \code{Real World} shows event occurrences over a continuous timeline 
divided in units of 10 milliseconds.
Esterel and \CEU use discrete logical units of time in which the occurring 
events (which are the same in all boxes) have to fit somehow:
%
\begin{itemize}
\item \code{[Box-1]}:
``Esterel with fixed-length ticks'',
within which reactions to occurring inputs have to fit~\cite{esterel.fixed}.
%
We assume a \code{R(boot)} reaction at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1}, causing a 
\emph{timing violation}~\cite{esterel.fixed} that invades \code{tick-2}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur ``simultaneously'' during the idle \code{tick-3}.
However, the program can only detect one occurrence of \code{E}, which is 
considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
``Esterel with variable-length ticks'', which are adjusted to the CPU time the 
corresponding reaction takes to complete~\cite{esterel.variable}.
This approach avoids the time violation for \code{R(A)} and also results in
smaller idle periods.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} on \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered on \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
``\CEU with unique and queue-based input events''.
We also assume a \code{R(boot)} reaction before any input.
%
Because the occurrence of event \code{A} is unique during \code{tick-0}, the 
behavior in \CEU is similar to \code{Box-2} for the first two units of time 
(\code{boot-0} and \code{A-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle period in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU recognizes and reacts to the two instances of \code{E} 
independently, which are handled in sequence.
\end{itemize}
